---
title: "NLR - CRP - BISAP"
author: "An Tran"
date: "2024-10-06"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE, fig.dpi=800,fig.path="figures/"}
knitr::opts_chunk$set(echo = F, warning = F, dpi=300, fig.width = 6)
library(randomForest)
library(gtsummary)
library(pROC)
library(rsample)
library(ggplot2)
library(ggsci)
library(rms)
library(dcurves)
library(survival)
library(tidyverse)
library(Hmisc)
library(readxl)
library(tidymodels)
library(rpart)
library(rpart.plot)


df1 <- read_excel("C:/Users/ADMIN/Desktop/Projects/SAP/data/data.xls", sheet = 2)
df1 <- as.data.frame(df1)
```

## Clean data

```{r}
df <- filter(df1, crp != 0)
df <- mutate(df, nlr = neu/lym)
df$phando2 <- factor(df$phando2, levels = c(0,1), labels = c(0,1))
df$sap <- factor(df$phando2, levels = c(0,1), labels = c("Non servere", "Severe"))
df$sirs <- factor(df$sirs, levels = c(0,1), labels = c(0,1))
df$tdmp <- as.factor(df$tdmp)
df$gioi <- as.factor(df$gioi)
df$nguyennhan <- as.factor(df$nguyennhan)
df$nn1 <- ifelse(df$nguyennhan==1,1,0)
df$nn2 <- ifelse(df$nguyennhan==2,1,0)
df$nn3 <- ifelse(df$nguyennhan==3,1,0)
df$nn4 <- ifelse(df$nguyennhan==4,1,0)
df$nn5 <- ifelse(df$nguyennhan==5,1,0)
df$nn1 <- as.factor(df$nn1)
df$nn2 <- as.factor(df$nn2)
df$nn3 <- as.factor(df$nn3)
df$nn4 <- as.factor(df$nn4)
df$nn5 <- as.factor(df$nn5)
```

## Devide data into 60% training set and 40% validation set

```{r}
set.seed(234)
samp <- sample(c(TRUE, FALSE), nrow(df), replace = TRUE, prob = c(0.6,0.4))
train <- df[samp, ]
test <- df[!samp, ]
```

```{r}
train$type <- "train"
test$type <- "test"
df <- full_join(train, test)
df$type <- as.factor(df$type)
```

## Descriptive summary

```{r}
tbl_summary(df, by = type, include = c(gioi, hct, tdmp, nguyennhan, tha,
                                       dtd, bisap, nanggia, tudich,
                                       hoaitutuy, hoaituwall,
                                       ketcuoc, phando2, sirs),
            missing = "no",
            digits = c(all_continuous() ~ 1,
                       all_categorical() ~ c(0,1)),
            statistic =  list(all_continuous() ~ "{mean} ({sd})")) %>%
  add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3),
                    test = list(all_continuous() ~ "t.test", 
                                all_categorical() ~ "fisher.test")) %>%
  add_overall()
```

```{r}
tbl_summary(df, by = type, include = c(tuoi, nlr, crp, lym, bun, cre, neu,
                                       lym, wbc, tri,lipase, amylase, ctsi),
            missing = "no",
             digits = c(all_continuous() ~ 1,
                         all_categorical() ~ c(0,1)),
              statistic =  list(all_continuous() ~ "{median} ({p25}-{p75})")) %>%
  add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3),
                    test = list(all_continuous() ~ "wilcox.test",
                                all_categorical() ~ "fisher.test")) %>%
  add_overall()
```

## Heatmap

```{r}
library(GGally)
df_pairs <- df %>%
  select(nlr, crp, bisap, phando2) %>%
  mutate(across(c(nlr, crp, bisap), as.numeric),
         phando2 = factor(phando2, levels = c(0, 1), labels = c("Non-Severe", "Severe")))

ggpairs(df_pairs,
        columns = 1:3,
        columnLabels = c("NLR", "CRP", "BISAP"),
        aes(color = phando2, alpha = 0.7),
        upper = list(continuous = wrap("cor", method = "spearman")),
        lower = list(continuous = "points"),
        diag = list(continuous = "densityDiag")) +
  theme(
    strip.text = element_text(size = 14, color = "black"),
    axis.text = element_text(size = 14, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    legend.text = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 14, color = "black")
  )
```

## Violin plot for CLR and BISAP

### Training data

```{r}
violin_nlr_train <- ggplot(train, aes(x=sap, y=nlr, fill=sap)) + 
  geom_violin(trim=FALSE) + scale_fill_manual(values = c( "#E69F00", "#56B4E9")) +
  geom_boxplot(width=0.1, fill="gray")+
  labs(x="Severity of Acute Pancreatitis", 
       y = "NLR value") +
  theme_classic() + 
  theme(legend.position = "none", 
        legend.title = element_text(face = "bold"))
```

``` {r}
violin_crp_train <- ggplot(train, aes(x=sap, y=crp, fill=sap)) + 
  geom_violin(trim=FALSE) + scale_fill_manual(values = c( "#E69F00", "#56B4E9")) +
  geom_boxplot(width=0.1, fill="gray")+
  labs(x="Severity of Acute Pancreatitis", 
       y = "CRP value") +
  theme_classic() + 
  theme(legend.position = "none", 
        legend.title = element_text(face = "bold"))
```

### Testing data

```{r}
violin_nlr_test <- ggplot(test, aes(x=sap, y=nlr, fill=sap)) + 
  geom_violin(trim=FALSE) + scale_fill_manual(values = c( "#E69F00", "#56B4E9")) +
  geom_boxplot(width=0.1, fill="gray")+
  labs(x="Severity of Acute Pancreatitis", 
       y = "NLR value") +
  theme_classic() + 
  theme(legend.position = "none", 
        legend.title = element_text(face = "bold"))
```

``` {r}
violin_crp_test <- ggplot(test, aes(x=sap, y=crp, fill=sap)) + 
  geom_violin(trim=FALSE) + scale_fill_manual(values = c( "#E69F00", "#56B4E9")) +
  geom_boxplot(width=0.1, fill="gray")+
  labs(x="Severity of Acute Pancreatitis", 
       y = "CRP value") +
  theme_classic() + 
  theme(legend.position = "none", 
        legend.title = element_text(face = "bold"))
```

## Build models

```{r}
train$phando2 <- factor(train$phando2, levels = c(0,1), labels = c("nonsap", "sap"))
test$phando2 <- factor(test$phando2, levels = c(0,1), labels = c("nonsap", "sap"))
library(caret)
kfold <- trainControl(method = "cv",
                      number = 10,
                      savePredictions = "all",
                      classProbs = TRUE) 

m_nlr <- train(phando2 ~ nlr, data = train,
               method = "glm", family = "binomial",
               trControl = kfold)

m_bisap <- train(phando2 ~ bisap, data = train, 
               method = "glm", family = "binomial",
               trControl = kfold)

m_crp <- train(phando2 ~ crp, data = train,
             method = "glm", family = "binomial",
               trControl = kfold)

m_nlr_crp <- train(phando2 ~ nlr + crp, data = train,
                 method = "glm", family = "binomial",
               trControl = kfold)

m_neu <- train(phando2 ~ neu, data = train,
             method = "glm", family = "binomial",
               trControl = kfold)

m_lym<- train(phando2 ~ lym, data = train,
             method = "glm", family = "binomial",
               trControl = kfold)
```

## Predict models

```{r}
p_nlr <- predict(m_nlr, test, type = "prob")
roc_nlr <- roc(test$phando2, p_nlr[,2])

p_bisap <- predict(m_bisap, test, type = "prob")
roc_bisap <- roc(test$phando2, p_bisap[,2])

p_crp <- predict(m_crp, test, type = "prob")
roc_crp <- roc(test$phando2, p_crp[,2])

p_nlr_crp <- predict(m_nlr_crp, test, type = "prob")
roc_nlr_crp <- roc(test$phando2, p_nlr_crp[,2])

p_neu <- predict(m_neu, test, type = "prob")
roc_neu <- roc(test$phando2, p_neu[,2])

p_lym <- predict(m_lym, test, type = "prob")
roc_lym <- roc(test$phando2, p_lym[,2])
```

## ROC Curves NLR, CRP and Combined

### Training cohort

```{r}
p_nlr_train <- predict(m_nlr, train, type = "prob")
roc_nlr_train <- roc(train$phando2, p_nlr_train[,2])

p_bisap_train <- predict(m_bisap, train, type = "prob")
roc_bisap_train <- roc(train$phando2, p_bisap_train[,2])

p_crp_train <- predict(m_crp, train, type = "prob")
roc_crp_train <- roc(train$phando2, p_crp_train[,2])

p_nlr_crp_train <- predict(m_nlr_crp, train, type = "prob")
roc_nlr_crp_train <- roc(train$phando2, p_nlr_crp_train[,2])
```

```{r}
size_text_roc <- 4
x_text_roc <- 0.65
y_text_roc <- 0.25
roc.list_train <- list(roc_nlr_train,roc_crp_train, roc_nlr_crp_train)
ggroc(roc.list_train, aes = "color", size = 1, linetype = "solid", legacy.axes = TRUE) +
    annotate("text", x = x_text_roc, y = y_text_roc, size = size_text_roc, hjust = 0, 
             color = "#1f77b4",
             label = paste0("NLR = ", sprintf("%.3f", roc_nlr_train$auc))) +
    annotate("text", x = x_text_roc, y = y_text_roc - 0.1, size = size_text_roc, hjust = 0, 
             color = "#ff7f0e",
             label = paste0("CRP = ", sprintf("%.3f", roc_crp_train$auc))) +
    annotate("text", x = x_text_roc, y = y_text_roc - 0.2, size = size_text_roc, hjust = 0, 
             color = "#d340a2",
             label = paste0("NLR + CRP = ", sprintf("%.3f", roc_nlr_crp_train$auc))) +
    geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
                 color = "darkgrey", linetype = "dashed", size = 0.8) +  # Đường y=x
    theme_classic() +
    theme(text = element_text(size = 14),
          legend.position = "bottom",
          legend.title = element_blank()) +
    scale_color_bmj(name = "Model", 
labels = c("NLR", "CRP", "NLR + CRP")) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    theme_minimal()
```

### Validation cohort

```{r}
size_text_roc <- 4
x_text_roc <- 0.65
y_text_roc <- 0.25
roc.list <- list(roc_nlr, roc_crp, roc_nlr_crp)
ggroc(roc.list, aes = "color", size = 1, linetype = "solid", legacy.axes = TRUE) +
    annotate("text", x = x_text_roc, y = y_text_roc, size = size_text_roc, hjust = 0, 
             color = "#1f77b4",
             label = paste0("NLR = ", sprintf("%.3f", roc_nlr$auc))) +
    annotate("text", x = x_text_roc, y = y_text_roc - 0.1, size = size_text_roc, hjust = 0, 
             color = "#ff7f0e",
             label = paste0("CRP = ", sprintf("%.3f", roc_crp$auc))) +
    annotate("text", x = x_text_roc, y = y_text_roc - 0.2, size = size_text_roc, hjust = 0, 
             color = "#d340a2",
             label = paste0("NLR + CRP = ", sprintf("%.3f", roc_nlr_crp$auc))) +
    geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
                 color = "darkgrey", linetype = "dashed", size = 0.8) +  # Đường y=x
    theme_classic() +
    theme(text = element_text(size = 14),
          legend.position = "bottom",
          legend.title = element_blank()) +
    scale_color_bmj(name = "Model", 
labels = c("NLR", "CRP", "NLR + CRP")) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    theme_minimal()
```

### Compare ROC Curve

```{r}
roc.test(roc_nlr, roc_crp)
roc.test(roc_nlr_crp, roc_nlr)
roc.test(roc_nlr_crp, roc_crp)
```

## Cut-off for NLR

### Training

```{r}
cpnlr_train <- cutpointr::cutpointr(data = train, x = nlr, class = phando2, metric = cutpointr::youden)
summary(cpnlr_train)
```

### Validation

```{r}
cpnlr_test <- cutpointr::cutpointr(data = test, x = nlr, class = phando2, 
                                   metric = cutpointr::youden)
summary(cpnlr_test)
```

## Cut-off for CRP

### Training

```{r}
cpcrp_train <- cutpointr::cutpointr(data = train, x = crp, class = phando2, 
                                    metric = cutpointr::youden)
summary(cpcrp_train)
```

### Validation

```{r}
cpcrp_test <- cutpointr::cutpointr(data = test, x = crp, class = phando2, 
                                   metric = cutpointr::youden)
summary(cpcrp_test)
```

## Decision tree for NLR and CRP

```{r}
set.seed(234)
tree_fit <- rpart(phando2 ~  nlr + crp, data = train,
              control = rpart.control(cp = 0.0001))

printcp(tree_fit)

best <- tree_fit$cptable[which.min(tree_fit$cptable[,"xerror"]),"CP"]

pruned_tree_fit <- prune(tree_fit, cp = best)

prp(pruned_tree_fit,
    faclen=0,
    extra=101,
    under = T,
    cex = 0.8,
    box.palette = "auto",
    roundint=F,
    digits=5)

predictions <- predict(tree_fit, test, type = "class")

confusionMatrix(predictions, test$phando2)
```

## Odds Ratio

### Training data

```{r}
train$phando2 <- ifelse(train$phando2=="nonsap", 0, 1)

dff <- train[,c("nlr", "crp", "phando2", "tuoi", "gioi", 
               "nguyennhan", "dtd", "wbc")]

dff <- dff %>% 
  mutate(
    nlr_or = ifelse(nlr>=11.4,1,0),
    crp_or = ifelse(crp>=174.7,1,0),
    combine_or = case_when(
      nlr < 11.4 ~ 1,
      nlr >= 11.4 & crp < 173.3 ~ 2,
      nlr >= 11.4 & crp >= 173.3 ~ 3)
  )

dff <- dff %>% mutate(across(.cols = -c(nlr, crp, tuoi, wbc), .fns = as.factor))

# Univariate 
tbl_uvregression(dff[,-c(1,2,4,5,6,7,8)],
                 method = glm,
                 y = phando2,
                 method.args = list(family = binomial),
                 exponentiate = TRUE)

## Multivariate - adjusted for age, sex, etiology, diabetes, wbc
glm(phando2 ~ nlr_or + tuoi + gioi + nguyennhan + dtd + wbc, data = dff, family = binomial(link = "logit")) %>%
  tbl_regression(exponentiate = TRUE)

glm(phando2 ~ crp_or + tuoi + gioi + nguyennhan + dtd + wbc, data = dff, family = binomial(link = "logit")) %>%
  tbl_regression(exponentiate = TRUE)

glm(phando2 ~ combine_or + tuoi + gioi + nguyennhan + dtd + wbc, data = dff, family = binomial(link = "logit")) %>%
  tbl_regression(exponentiate = TRUE)
```

### Testing data

```{r}
test$phando2 <- ifelse(test$phando2=="nonsap", 0, 1)

dff <- test[,c("nlr", "crp", "phando2", "tuoi", "gioi", 
               "nguyennhan", "dtd", "wbc")]

dff <- dff %>% 
  mutate(
    nlr_or = ifelse(nlr>=11.43,1,0),
    crp_or = ifelse(crp>=174.7,1,0),
    combine_or = case_when(
      nlr < 11.4 ~ 1,
      nlr >= 11.4 & crp < 173.3 ~ 2,
      nlr >= 11.4 & crp >= 173.3 ~ 3)
  )

dff$phando2 <- ifelse(dff$phando2=="nonsap", 0, 1)

# Univariate 
tbl_uvregression(dff[,-c(1,2,3,4,5,6,7))],
                 method = glm,
                 y = phando2,
                 method.args = list(family = binomial),
                 exponentiate = TRUE)

## Multivariate - adjusted for age, sex, etiology, diabetes, wbc
glm(phando2 ~ nlr_or + tuoi + gioi + nguyennhan + dtd + wbc, data = dff, family = binomial(link = "logit")) %>%
  tbl_regression(exponentiate = TRUE)

glm(phando2 ~ crp_or + tuoi + gioi + nguyennhan + dtd + wbc, data = dff, family = binomial(link = "logit")) %>%
  tbl_regression(exponentiate = TRUE)

glm(phando2 ~ combine_or + tuoi + gioi + nguyennhan + dtd + wbc, data = dff, family = binomial(link = "logit")) %>%
  tbl_regression(exponentiate = TRUE)
```

## ROC Curves NLR + CRP and BISAP

### Training cohort

```{r}
size_text_roc <- 4
x_text_roc <- 0.65
y_text_roc <- 0.25
roc.list_train <- list(roc_nlr_crp_train, roc_bisap_train)
ggroc(roc.list_train, aes = "color", size = 1, linetype = "solid", legacy.axes = TRUE) +
    annotate("text", x = x_text_roc, y = y_text_roc, size = size_text_roc, hjust = 0, 
             color = "#1f77b4",
             label = paste0("NLR + CRP = ", sprintf("%.3f", roc_nlr_crp_train$auc))) +
    annotate("text", x = x_text_roc, y = y_text_roc - 0.1, size = size_text_roc, hjust = 0, 
             color = "#ff7f0e",
             label = paste0("BISAP = ", sprintf("%.3f", roc_bisap_train$auc))) +
    geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
                 color = "darkgrey", linetype = "dashed", size = 0.8) +  # Đường y=x
    theme_classic() +
    theme(text = element_text(size = 14),
          legend.position = "bottom",
          legend.title = element_blank()) +
    scale_color_bmj(name = "Model", 
labels = c("NLR + CRP", "BISAP")) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    theme_minimal()
```

### Validation cohort

```{r}
size_text_roc <- 4
x_text_roc <- 0.65
y_text_roc <- 0.25
roc.list <- list(roc_nlr_crp, roc_bisap)

ggroc(roc.list, aes = "color", size = 1, linetype = "solid", legacy.axes = TRUE) +
    annotate("text", x = x_text_roc, y = y_text_roc, size = size_text_roc, hjust = 0, 
             color = "#1f77b4",
             label = paste0("NLR + CRP = ", sprintf("%.3f", roc_nlr_crp$auc))) +
    annotate("text", x = x_text_roc, y = y_text_roc - 0.1, size = size_text_roc, hjust = 0, 
             color = "#ff7f0e",
             label = paste0("BISAP = ", sprintf("%.3f", roc_bisap$auc))) +
  annotate("text", x = x_text_roc, y = y_text_roc - 0.2, size = size_text_roc, hjust = 0, 
             color = "#ff7f0e", 
           label = paste0("(p = 0.286)")) +
    geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
                 color = "darkgrey", linetype = "dashed", size = 0.8) +  # Đường y=x
    theme_classic() +
    theme(text = element_text(size = 14),
          legend.position = "bottom",
          legend.title = element_blank()) +
    scale_color_bmj(name = "Model", 
labels = c("NLR + CRP", "BISAP")) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    theme_minimal()
```

```{r}
roc.test(roc_nlr_crp, roc_bisap)
```

## Calibration NLR + CRP

### Training set

```{r}
df_plot <- train[,c("nlr", "crp","phando2")]
df_plot <- na.omit(df_plot)
pred.train <- predict(m_nlr_crp, df_plot, type = "prob")
phat.train <- pred.train[,2]
val.prob(phat.train, as.numeric(df_plot$phando2) - 1, m=20, cex=.5)
```

### Validation set

```{r}
df_plot <- test[,c("nlr", "crp","phando2")]
df_plot <- na.omit(df_plot)
pred.test <- predict(m_nlr_crp, df_plot, type = "prob")
phat.test <- pred.test[,2]
val.prob(phat.test, as.numeric(df_plot$phando2) - 1, m=20, cex=.5)
```

## Decision curve NLR + CRP

### Training set

```{r}
df_plot <- train[,c("phando2","nlr","crp","bisap")]
df_plot <- na.omit(df_plot)
pred1 <- predict(m_nlr_crp, newdata = df_plot, type = 'prob')
pred2 <- predict(m_bisap, newdata = df_plot, type = 'prob')
df_plot$pred1 <- pred1[,2]
df_plot$pred2 <- pred2[,2]
df_plot$phando2 <- as.numeric(df_plot$phando2) - 1
dca(phando2 ~ pred1 + pred2,
    data = df_plot,
    label = list(pred1 = "NLR + CRP",
                 pred2 = "BISAP"))
```

### Validation set

```{r}
df_plot <- test[,c("phando2","nlr","crp", "bisap")]
df_plot <- na.omit(df_plot)
pred1 <- predict(m_nlr_crp, newdata = df_plot, type = 'prob')
pred2 <- predict(m_bisap, newdata = df_plot, type = 'prob')
df_plot$pred1 <- pred1[,2]
df_plot$pred2 <- pred2[,2]
df_plot$phando2 <- as.numeric(df_plot$phando2) - 1
dca(phando2 ~ pred1 + pred2,
    data = df_plot,
    label = list(pred1 = "NLR + CRP",
                 pred2 = "BISAP"))
```
